---
title: "Cobos-Uribe et al 2024: Clinical Data Analysis"
author: "Catalina Cobos-Uribe"
date: "2024-12-18"
output: html_document
---

# Purpose
This script analyses the inflammatory cell and cytokine data from the study by Cobos-Uribe et al. (2024). This study aimed to investigate the effect of controlled wood smoke exposure on the respiratory microbiome and immune response. The analysis includes data processing, statistical tests and visualization of key findings.

# Input:
"data/metadada.csv"

#### Load libraries
```{r libraries, include=FALSE }
library(rstatix)
library(ggpubr)
library(ggplot2)
library(gridExtra) 
library(grid)
library(dplyr)
library(grid)
```

#### Prepare data
```{r}
data <- read.csv("metadata.csv")

clinical_data <- data %>% 
  filter(sample_type == "sputum") 

cells <- clinical_data %>%
  filter(!is.na(clinical_data$sputum_mac_percentage) & !is.na(clinical_data$sputum_PMN_percentage) & !is.na(clinical_data$sputum_macs_cells.mg) & !is.na(clinical_data$sputum_pmn_cells.mg) &
          sputum_mac_percentage > 0 &
           sputum_PMN_percentage > 0 &
           sputum_macs_cells.mg >0 &
           sputum_pmn_cells.mg > 0)

cells <- cells %>% 
  filter(!subject %in% c("6", "7", "11",
                         "28", "70", "72", "75", "79"))

cells <- cells %>% 
  mutate(
    log_pmn_mg = log1p(sputum_pmn_cells.mg),
    log_mac_mg = log1p(sputum_macs_cells.mg),
    log_total_mg = log1p(sputum._cells.mg)
  )

cyto <- clinical_data %>%
  filter(!is.na(clinical_data$sputum_il1b) & !is.na(clinical_data$sputum_il6) & !is.na(clinical_data$sputum_il8) & !is.na(clinical_data$sputum_TNFa) &
          sputum_il1b > 0 &
           sputum_il6 > 0 &
           sputum_il8 >0 &
           sputum_TNFa > 0)

cyto <- cyto %>% 
  filter(!subject %in% c("41", "46", "60",
                         "62"))
cyto <- cyto %>% 
  mutate(
    log_il1b = log1p(sputum_il1b),
    log_il6 = log1p(sputum_il6),
    log_il8 = log1p(sputum_il8),
    log_tnfa = log1p(sputum_TNFa)
  )

```

#### Prepare for plots
```{r}
# Create new x-axis labels
new_labels <- c("Pre-exposure", "6h post-exposure", "24h post-exposure")
clinical_data$time <- factor(clinical_data$time, levels = c("train", "post", "fu"))
cells$time <- factor(cells$time, levels = c("train", "post", "fu"))
cyto$time <- factor(cyto$time, levels = c("train", "post", "fu"))

```

#### Assess normality of inflmmatory cells and cytokines
```{r}
# Assess normality: QQ-plot

png("qq_plots/pmn_per_qq.png", width = 6, height = 4, units = "in", res = 300)

par(mfrow = c(1, 3))
for (t in levels(cells$time)) {
  subset <- cells$sputum_PMN_percentage[cells$time == t]
  plot_title <- switch(t,
                       "train" = "Neutrophil%: Pre-Exposure",
                       "post" = "6h Post-Exposure",
                       "fu" = "24h Post-Exposure")
  qqnorm(subset, main = plot_title)
  qqline(subset, col = "red")
}

dev.off()


png("qq_plots/mac_mg_qq.png", width = 6, height = 4, units = "in", res = 300)

par(mfrow = c(1, 3))
for (t in levels(cells$time)) {
  subset <- cells$sputum_macs_cells.mg[cells$time == t]
  log_subset <- log1p(subset)
  plot_title <- switch(t,
                       "train" = "Macrophage/mg: Pre-Exposure",
                       "post" = "6h Post-Exposure",
                       "fu" = "24h Post-Exposure")
  qqnorm(log_subset, main = plot_title)
  qqline(log_subset, col = "red")
}

dev.off()


png("qq_plots/il8_qq.png", width = 6, height = 4, units = "in", res = 300)

par(mfrow = c(1, 3))
for (t in levels(cyto$time)) {
  subset <- cyto$sputum_il8[cyto$time == t]
  log_subset <- log1p(subset)
  plot_title <- switch(t,
                       "train" = "IL-8: Pre-Exposure",
                       "post" = "6h Post-Exposure",
                       "fu" = "24h Post-Exposure")
  qqnorm(log_subset, main = plot_title)
  qqline(log_subset, col = "red")
}

dev.off()
```


### Macrophage% analysis and visualization
```{r Mac% - one-way repeated measures ANOVA}
res.aov_mac <- anova_test(data = cells, dv=sputum_mac_percentage, 
             wid=subject,
             within = time)
get_anova_table(res.aov_mac)

pwc_mac <- cells %>% 
  pairwise_t_test(
    sputum_mac_percentage ~ time, paired = TRUE,
    p.adjust.method = "bonferroni"
    )
pwc_mac


p3 <- ggplot(data = cells, 
       aes(x = time, y = sputum_mac_percentage, group = subject, color)) +
  geom_point(size = 3, show.legend = FALSE) +              
  geom_line(alpha = 0.6, show.legend = FALSE) +            
  stat_summary(fun = mean, geom = "line", aes(group = 1), 
               color = "blue", linewidth = 1.2, linetype = "dashed") +  
  stat_summary(fun = mean, geom = "point", aes(group = 1), 
               color = "blue", size = 3) +  
  scale_x_discrete(labels = new_labels) +   
  ylab("Macrophage %") +                    
  xlab(NULL) +                              
  theme_minimal()  

pwc_mac <- pwc_mac %>% add_xy_position(x = "time")
p3 <- p3 + 
  stat_pvalue_manual(pwc_mac, hide.ns = TRUE, step.increase = 0.1,
                     y.position = 100)+
  labs(subtitle = NULL, caption = NULL)

p3
```
```{r Mac% - two-way repeated measures ANOVA}
res.aov_mac <- anova_test(
  data = cells, 
  dv=sputum_mac_percentage, 
             wid=subject,
             within = time,
  between = sex)
get_anova_table(res.aov_mac)

# SEX EFFECT 
pwc <- cells %>%
  group_by(time) %>%
  pairwise_t_test(
    sputum_mac_percentage ~ sex, paired = FALSE,
    p.adjust.method = "bonferroni"
    )
pwc

# EFFECT OF TIME
pwc2 <- cells %>%
  group_by(sex) %>%
  pairwise_t_test(
    sputum_mac_percentage ~ time, paired = TRUE,
    p.adjust.method = "bonferroni"
    )
pwc2

p7 <- ggboxplot(cells, x = "time", y = "sputum_mac_percentage", add = "jitter", 
                 color = "sex",
                 palette = c("purple", "turquoise"), 
                 ylab = "Macrophage %",
                 xlab = FALSE,
                 bxp.errorbar = TRUE)+
  scale_x_discrete(labels= new_labels)+
  theme(legend.title = element_blank())


pwc2 <- pwc2 %>% add_xy_position(x = "time")
pwc <- pwc %>% add_xy_position(x = "time")
p7 <- p7 + 
  stat_pvalue_manual(pwc2, hide.ns = TRUE, step.increase = 0.1,
                     y.position = 100,
                     color = "purple")+
  labs(subtitle = NULL, caption = NULL)

p7 <- p7 + 
  stat_pvalue_manual(pwc, hide.ns = TRUE, step.increase = 0.2,
                     y.position = 120)+
  labs(subtitle = NULL, caption = NULL)
p7
```


### PMN% analysis and plot
```{r one-way repeated measures ANOVA}
res.aov_pmn <- anova_test(data = cells, dv=sputum_PMN_percentage, 
             wid=subject,
             within = time)
get_anova_table(res.aov_pmn)

pwc_pmn <- cells %>% 
  pairwise_t_test(
    sputum_PMN_percentage ~ time, paired = TRUE,
    p.adjust.method = "bonferroni"
    )
pwc_pmn


p1 <- ggplot(data = cells, 
       aes(x = time, y = sputum_PMN_percentage, group = subject)) +
  geom_point(size = 3) +              
  geom_line(alpha = 0.6) +            
  stat_summary(fun = mean, geom = "line", aes(group = 1), 
               color = "blue", linewidth = 1.2, linetype = "dashed") +  
  stat_summary(fun = mean, geom = "point", aes(group = 1), 
               color = "blue", size = 3) +  
  scale_x_discrete(labels = new_labels) +   
  ylab("Neutrophil %") +                    
  xlab(NULL) +                              
  theme_minimal() 

pwc_pmn <- pwc_pmn %>% add_xy_position(x = "time")
p1 <- p1 + 
  stat_pvalue_manual(pwc_pmn, hide.ns = TRUE, step.increase = 0.1, y.position = 100) +
  labs(
    subtitle = NULL, caption = NULL
  )

p1
```
```{r PMN% - two-way repeated measures ANOVA}
res.aov_pmn <- anova_test(
  data = cells, 
  dv=sputum_PMN_percentage, 
             wid=subject,
             within = time,
  between = sex)
get_anova_table(res.aov_pmn)

# SEX EFFECT 
pwc <- cells %>%
  group_by(time) %>%
  pairwise_t_test(
    sputum_PMN_percentage ~ sex, paired = FALSE,
    p.adjust.method = "bonferroni"
    )
pwc

# EFFECT OF TIME
pwc2 <- cells %>%
  group_by(sex) %>%
  pairwise_t_test(
    sputum_PMN_percentage ~ time, paired = TRUE,
    p.adjust.method = "bonferroni"
    )
pwc2

p5 <- ggboxplot(cells, x = "time", y = "sputum_PMN_percentage", add = "jitter", 
                 color = "sex",
                 palette = c("purple", "turquoise"), 
                 ylab = "Neutrophil %",
                 xlab = FALSE,
                 bxp.errorbar = TRUE)+
  scale_x_discrete(labels= new_labels)+
  theme(legend.title = element_blank())


pwc2 <- pwc2 %>% add_xy_position(x = "time")
pwc <- pwc %>% add_xy_position(x = "time")

p5 <- p5 + 
  stat_pvalue_manual(pwc2, hide.ns = TRUE, step.increase = 0.1,
                     y.position = 100,
                     color = "purple")+
  labs(subtitle = NULL, caption = NULL)

p5 <- p5 + 
  stat_pvalue_manual(pwc, hide.ns = TRUE, step.increase = 0.2,
                     y.position = 120)+
  labs(subtitle = NULL, caption = NULL)
p5
```

### Macrophages per mg
```{r}
res.aov_mac_mg <- anova_test(data = cells, dv=sputum_macs_cells.mg, 
             wid=subject,
             within = time)
get_anova_table(res.aov_mac_mg)

pwc_mac_mg <- cells %>% 
  pairwise_t_test(
    sputum_macs_cells.mg ~ time, paired = TRUE,
    p.adjust.method = "bonferroni"
    )
pwc_mac_mg

#absolute macs after log transformation
res.aov_mac_mg <- anova_test(data = cells, dv = log_mac_mg, 
             wid=subject,
             within = time)
get_anova_table(res.aov_mac_mg)

pwc_mac_mg <- cells %>% 
  pairwise_t_test(
    log_mac_mg ~ time, paired = TRUE,
    p.adjust.method = "bonferroni"
    )
pwc_mac_mg


p4 <- ggplot(data = cells, 
       aes(x = time, y = sputum_macs_cells.mg, group = subject)) +
  geom_point(size = 3, show.legend =  FALSE) +              
  geom_line(alpha = 0.6, show.legend = FALSE) +            
  stat_summary(fun = mean, geom = "line", aes(group = 1), 
               color = "blue", linewidth = 1.2, linetype = "dashed") +  
  stat_summary(fun = mean, geom = "point", aes(group = 1), 
               color = "blue", size = 3) +  
  scale_x_discrete(labels = new_labels) +  
  ylab("Macrophages per sputum mg") +                    
  xlab(NULL) +                              
  theme_minimal() +
  scale_y_continuous(trans = "log10", 
                     breaks = c(1, 10, 100, 1000, 10000),
                     labels = c("1", "10", "100", "1,000", "10,000"))

pwc_mac_mg <- pwc_mac_mg %>% add_xy_position(x = "time")
p4 <- p4 + 
  stat_pvalue_manual(pwc_mac_mg, hide.ns = TRUE, y.position = 3.8)+
  labs(subtitle = NULL, caption = NULL)

p4
```
```{r Macs/mg - two-way repeated measures ANOVA}
res.aov_mac <- anova_test(
  data = cells, 
  dv=log_mac_mg, 
             wid=subject,
             within = time,
  between = sex)
get_anova_table(res.aov_mac)

# SEX EFFECT 
pwc <- cells %>%
  group_by(time) %>%
  pairwise_t_test(
    log_mac_mg ~ sex, paired = FALSE,
    p.adjust.method = "bonferroni"
    )
pwc

# EFFECT OF TIME
pwc2 <- cells %>%
  group_by(sex) %>%
  pairwise_t_test(
    log_mac_mg ~ time, paired = TRUE,
    p.adjust.method = "bonferroni"
    )
pwc2

p8 <- ggboxplot(cells, x = "time", y = "sputum_macs_cells.mg", add = "jitter", 
                 color = "sex",
                 palette = c("purple", "turquoise"), 
                 ylab = "Macrophage per Sputum mg",
                 xlab = FALSE,
                 bxp.errorbar = TRUE)+
  scale_x_discrete(labels= new_labels)+
  theme(legend.title = element_blank())+
  scale_y_continuous(trans = "log10", 
                     breaks = c(1, 10, 100, 1000, 10000),
                     labels = c("1", "10", "100", "1,000", "10,000"))


pwc2 <- pwc2 %>% add_xy_position(x = "time")
pwc <- pwc %>% add_xy_position(x = "time")

p8 <- p8 + 
  stat_pvalue_manual(pwc2, hide.ns = TRUE, step.increase = 0.1,
                     y.position = 100,
                     color = "purple")+
  labs(subtitle = NULL, caption = NULL)

p8 <- p8 + 
  stat_pvalue_manual(pwc, hide.ns = TRUE, step.increase = 0.2,
                     y.position = 120)+
  labs(subtitle = NULL, caption = NULL)
p8
```

### Neutrophils per mg
```{r PMN/mg - one-way repeated measures ANOVA}
res.aov_pmn_mg <- anova_test(data = cells, dv=sputum_pmn_cells.mg, 
             wid=subject,
             within = time)
get_anova_table(res.aov_pmn_mg)


#Log transform absolute neutrophils
res.aov_pmn_mg <- anova_test(data = cells, dv= log_pmn_mg, 
             wid=subject,
             within = time)
get_anova_table(res.aov_pmn_mg)



p2 <- ggplot(data = cells, 
       aes(x = time, y = sputum_pmn_cells.mg, group = subject)) +
  geom_point(size = 3, show.legend =  FALSE) +              
  geom_line(alpha = 0.6, show.legend = FALSE) +            
  stat_summary(fun = mean, geom = "line", aes(group = 1), 
               color = "blue", linewidth = 1.2, linetype = "dashed") +  
  stat_summary(fun = mean, geom = "point", aes(group = 1), 
               color = "blue", size = 3) +  
  scale_x_discrete(labels = new_labels) +   
  ylab("Neutrophils per sputum mg") +                    
  xlab(NULL) +                              
  theme_minimal() +
  scale_y_continuous(trans = "log10")

p2
```
```{r PMN/mg - two-way repeated measures ANOVA}

res.aov_pmn <- anova_test(
  data = cells, 
  dv=log_pmn_mg, 
             wid=subject,
             within = time,
  between = sex)
get_anova_table(res.aov_pmn)

# SEX EFFECT 
pwc <- cells %>%
  group_by(time) %>%
  pairwise_t_test(
    log_pmn_mg ~ sex, paired = FALSE,
    p.adjust.method = "bonferroni"
    )
pwc

# EFFECT OF TIME
pwc2 <- cells %>%
  group_by(sex) %>%
  pairwise_t_test(
    log_pmn_mg ~ time, paired = TRUE,
    p.adjust.method = "bonferroni"
    )
pwc2

p6 <- ggboxplot(cells, x = "time", y = "sputum_pmn_cells.mg", add = "jitter", 
                 color = "sex",
                 palette = c("purple", "turquoise"), 
                 ylab = "Neutrophils per Sputum mg",
                 xlab = FALSE,
                 bxp.errorbar = TRUE)+
  scale_x_discrete(labels= new_labels)+
  theme(legend.title = element_blank())+
  scale_y_continuous(trans = "log10", 
                     breaks = c(1, 10, 100, 1000),
                     labels = c("1", "10", "100", "1,000"))

pwc <- pwc %>% add_xy_position(x = "time")

p6 <- p6 + 
  stat_pvalue_manual(pwc, hide.ns = TRUE)+
  labs(subtitle = NULL, caption = NULL)
p6
```

### Combine plots
```{r}
combined <- grid.arrange(p1, p2, p3, p4, ncol = 2)
png("Inflammatory_cells_dot_line_plot_log.png", width = 13, height = 8, units = "in", res = 300)
grid.draw(combined)
grid.text("A", x = unit(0.01, "npc"), y = unit(0.95, "npc"), just = "left", gp = gpar(fontsize = 14, fontface = "bold"))
grid.text("B", x = unit(0.5, "npc"), y = unit(0.95, "npc"), just = "left", gp = gpar(fontsize = 14, fontface = "bold"))
grid.text("C", x = unit(0.01, "npc"), y = unit(0.48, "npc"), just = "left", gp = gpar(fontsize = 14, fontface = "bold"))
grid.text("D", x = unit(0.5, "npc"), y = unit(0.48, "npc"), just = "left", gp = gpar(fontsize = 14, fontface = "bold"))

dev.off()
```
### Combine plots by sex
```{r}
combined <- grid.arrange(p5, p6, p7, p8, ncol = 2)
png("clinical_supp_figures/Inflammatory_cells_bySex.png", width = 13, height = 8, units = "in", res = 300)
grid.draw(combined)
grid.text("A", x = unit(0.01, "npc"), y = unit(0.95, "npc"), just = "left", gp = gpar(fontsize = 14, fontface = "bold"))
grid.text("B", x = unit(0.5, "npc"), y = unit(0.95, "npc"), just = "left", gp = gpar(fontsize = 14, fontface = "bold"))
grid.text("C", x = unit(0.01, "npc"), y = unit(0.48, "npc"), just = "left", gp = gpar(fontsize = 14, fontface = "bold"))
grid.text("D", x = unit(0.5, "npc"), y = unit(0.48, "npc"), just = "left", gp = gpar(fontsize = 14, fontface = "bold"))

dev.off()
```

## Total cells per mg
```{r}
res.aov_total <- anova_test(data = cells, dv=sputum._cells.mg, 
             wid=subject,
             within = time)
get_anova_table(res.aov_total) # ns

# Log transformed data
res.aov_total <- anova_test(data = cells, dv=log_total_mg, 
             wid=subject,
             within = time)
get_anova_table(res.aov_total) #ns

p <- ggplot(data = cells, 
       aes(x = time, y = sputum._cells.mg, group = subject)) +
  geom_point(size = 3, show.legend =  FALSE) +              
  geom_line(alpha = 0.6, show.legend = FALSE) +            
  stat_summary(fun = mean, geom = "line", aes(group = 1), 
               color = "blue", linewidth = 1.2, linetype = "dashed") +  
  stat_summary(fun = mean, geom = "point", aes(group = 1), 
               color = "blue", size = 3) +  
  scale_x_discrete(labels = new_labels) +   
  ylab("Total Cells per Sputum mg") +                    
  xlab(NULL) +                              
  theme_classic () +
  scale_y_continuous(trans = "log10")

p

ggsave("clinical_supp_figures/total_cells_mg.png", plot = p, width = 6, height = 4, dpi = 300)
```
#### Cytokine analysis
```{r IL-1 beta}
res.aov_il1b <- anova_test(data = cyto, dv=log_il1b, 
             wid=subject,
             within = time)
get_anova_table(res.aov_il1b)


p1 <- ggplot(data = cyto, 
       aes(x = time, y = sputum_il1b, group = subject)) +
  geom_point(size = 3, show.legend =  FALSE) +              
  geom_line(alpha = 0.6, show.legend = FALSE) +            
  stat_summary(fun = mean, geom = "line", aes(group = 1), 
               color = "blue", linewidth = 1.2, linetype = "dashed") +  
  stat_summary(fun = mean, geom = "point", aes(group = 1), 
               color = "blue", size = 3) +  
  scale_x_discrete(labels = new_labels) +   
  ylab("IL-1 beta (pg/ml)") +                    
  xlab(NULL) +                              
  theme_minimal() +
  scale_y_continuous(trans = "log10")

p1

```
```{r IL-6}
res.aov_il6 <- anova_test(data = cyto, dv=log_il6, 
             wid=subject,
             within = time)
get_anova_table(res.aov_il6)


p2 <- ggplot(data = cyto, 
       aes(x = time, y = sputum_il6, group = subject)) +
  geom_point(size = 3, show.legend =  FALSE) +              
  geom_line(alpha = 0.6, show.legend = FALSE) +            
  stat_summary(fun = mean, geom = "line", aes(group = 1), 
               color = "blue", linewidth = 1.2, linetype = "dashed") +  
  stat_summary(fun = mean, geom = "point", aes(group = 1), 
               color = "blue", size = 3) +  
  scale_x_discrete(labels = new_labels) +   
  ylab("IL-6 (pg/ml)") +                    
  xlab(NULL) +                              
  theme_minimal() +
  scale_y_continuous(trans = "log10")

p2

```
```{r IL-8}
res.aov_il8 <- anova_test(data = cyto, dv=log_il8, 
             wid=subject,
             within = time)
get_anova_table(res.aov_il8)


p3 <- ggplot(data = cyto, 
       aes(x = time, y = sputum_il8, group = subject)) +
  geom_point(size = 3, show.legend =  FALSE) +              
  geom_line(alpha = 0.6, show.legend = FALSE) +            
  stat_summary(fun = mean, geom = "line", aes(group = 1), 
               color = "blue", linewidth = 1.2, linetype = "dashed") +  
  stat_summary(fun = mean, geom = "point", aes(group = 1), 
               color = "blue", size = 3) +  
  scale_x_discrete(labels = new_labels) +   
  ylab("IL-8 (pg/ml)") +                    
  xlab(NULL) +                              
  theme_minimal() +
  scale_y_continuous(trans = "log10")

p3

```
```{r TNF alpha}
res.aov_tnfa <- anova_test(data = cyto, dv=log_tnfa, 
             wid=subject,
             within = time)
get_anova_table(res.aov_tnfa)


p4 <- ggplot(data = cyto, 
       aes(x = time, y = sputum_TNFa, group = subject)) +
  geom_point(size = 3, show.legend =  FALSE) +              
  geom_line(alpha = 0.6, show.legend = FALSE) +            
  stat_summary(fun = mean, geom = "line", aes(group = 1), 
               color = "blue", linewidth = 1.2, linetype = "dashed") +  
  stat_summary(fun = mean, geom = "point", aes(group = 1), 
               color = "blue", size = 3) +  
  scale_x_discrete(labels = new_labels) +   
  ylab("TNF alpha (pg/ml)") +                    
  xlab(NULL) +                              
  theme_minimal() +
  scale_y_continuous(trans = "log10")

p4

```
```{r combine cyto data}
combined <- grid.arrange(p1, p2, p3, p4, ncol = 2)

png("clinical_supp_figures/cyto.png", width = 13, height = 8, units = "in", res = 300)
grid.draw(combined)
grid.text("A", x = unit(0.01, "npc"), y = unit(0.95, "npc"), just = "left", gp = gpar(fontsize = 14, fontface = "bold"))
grid.text("B", x = unit(0.5, "npc"), y = unit(0.95, "npc"), just = "left", gp = gpar(fontsize = 14, fontface = "bold"))
grid.text("C", x = unit(0.01, "npc"), y = unit(0.48, "npc"), just = "left", gp = gpar(fontsize = 14, fontface = "bold"))
grid.text("D", x = unit(0.5, "npc"), y = unit(0.48, "npc"), just = "left", gp = gpar(fontsize = 14, fontface = "bold"))

dev.off()
```
```{r il-1 beta - two-way repeated measures ANOVA}
res.aov <- anova_test(
  data = cyto, 
  dv=log_il1b, 
             wid=subject,
             within = time,
  between = sex)
get_anova_table(res.aov) #ns

# SEX EFFECT 
# pwc <- cyto %>%
#   group_by(time) %>%
#   pairwise_t_test(
#     log_il1b ~ sex, paired = FALSE,
#     p.adjust.method = "bonferroni"
#     )
# pwc
# 
# # EFFECT OF TIME
# pwc2 <- cyto %>%
#   group_by(sex) %>%
#   pairwise_t_test(
#     log_il1b ~ time, paired = TRUE,
#     p.adjust.method = "bonferroni"
#     )
# pwc2

p5 <- ggboxplot(cyto, x = "time", y = "sputum_il1b", add = "jitter", 
                 color = "sex",
                 palette = c("purple", "turquoise"), 
                 ylab = "IL-1 beta (pg/ml)",
                 xlab = FALSE,
                 bxp.errorbar = TRUE)+
  scale_x_discrete(labels= new_labels)+
  theme(legend.title = element_blank())+
  scale_y_continuous(trans = "log10")


# pwc2 <- pwc2 %>% add_xy_position(x = "time")
# pwc <- pwc %>% add_xy_position(x = "time")
# 
# p5 <- p5 + 
#   stat_pvalue_manual(pwc2, hide.ns = TRUE, step.increase = 0.1,
#                      y.position = 100,
#                      color = "purple")+
#   labs(subtitle = NULL, caption = NULL)
# 
# p5 <- p5 + 
#   stat_pvalue_manual(pwc, hide.ns = TRUE, step.increase = 0.2,
#                      y.position = 120)+
#   labs(subtitle = NULL, caption = NULL)
# p5
```
```{r il-6 - two-way repeated measures ANOVA}
res.aov <- anova_test(
  data = cyto, 
  dv=log_il6, 
             wid=subject,
             within = time,
  between = sex)
get_anova_table(res.aov) 

#SEX EFFECT
pwc <- cyto %>%
  group_by(time) %>%
  pairwise_t_test(
    log_il6 ~ sex, paired = FALSE,
    p.adjust.method = "bonferroni"
    )
pwc
 
# EFFECT OF TIME
pwc2 <- cyto %>%
  group_by(sex) %>%
  pairwise_t_test(
    log_il6 ~ time, paired = TRUE,
    p.adjust.method = "bonferroni"
    )
pwc2

p6 <- ggboxplot(cyto, x = "time", y = "sputum_il6", add = "jitter", 
                 color = "sex",
                 palette = c("purple", "turquoise"), 
                 ylab = "IL-6 (pg/ml)",
                 xlab = FALSE,
                 bxp.errorbar = TRUE)+
  scale_x_discrete(labels= new_labels)+
  theme(legend.title = element_blank())+
  scale_y_continuous(trans = "log10", 
                     breaks = c(1, 10, 100),
                     labels = c("1", "10", "100"))


# pwc2 <- pwc2 %>% add_xy_position(x = "time")
pwc <- pwc %>% add_xy_position(x = "time")
 
# p5 <- p5 + 
#   stat_pvalue_manual(pwc2, hide.ns = TRUE, step.increase = 0.1,
#                      y.position = 100,
#                      color = "purple")+
#   labs(subtitle = NULL, caption = NULL)
# 
p6 <- p6 +
  stat_pvalue_manual(pwc, hide.ns = TRUE, step.increase = 0.2,
                     y.position = 3)+
  labs(subtitle = NULL, caption = NULL)
p6
```

```{r il-8 - two-way repeated measures ANOVA}
res.aov <- anova_test(
  data = cyto, 
  dv=log_il8, 
             wid=subject,
             within = time,
  between = sex)
get_anova_table(res.aov) 

#SEX EFFECT
pwc <- cyto %>%
  group_by(time) %>%
  pairwise_t_test(
    log_il8 ~ sex, paired = FALSE,
    p.adjust.method = "bonferroni"
    )
pwc
 
# EFFECT OF TIME
pwc2 <- cyto %>%
  group_by(sex) %>%
  pairwise_t_test(
    log_il8 ~ time, paired = TRUE,
    p.adjust.method = "bonferroni"
    )
pwc2

p7 <- ggboxplot(cyto, x = "time", y = "sputum_il8", add = "jitter", 
                 color = "sex",
                 palette = c("purple", "turquoise"), 
                 ylab = "IL-8 (pg/ml)",
                 xlab = FALSE,
                 bxp.errorbar = TRUE)+
  scale_x_discrete(labels= new_labels)+
  theme(legend.title = element_blank())+
  scale_y_continuous(trans = "log10", 
                     breaks = c(1, 10, 100, 1000),
                     labels = c("1", "10", "100", "1000"))


# pwc2 <- pwc2 %>% add_xy_position(x = "time")
pwc <- pwc %>% add_xy_position(x = "time")
 
# p7 <- p7 + 
#   stat_pvalue_manual(pwc2, hide.ns = TRUE, step.increase = 0.1,
#                      y.position = 100,
#                      color = "purple")+
#   labs(subtitle = NULL, caption = NULL)
# 
p7 <- p7 +
  stat_pvalue_manual(pwc, hide.ns = TRUE, step.increase = 0.2,
                     y.position = 4)+
  labs(subtitle = NULL, caption = NULL)
p7
```

```{r TNFa - two-way repeated measures ANOVA}
res.aov <- anova_test(
  data = cyto, 
  dv=log_tnfa, 
             wid=subject,
             within = time,
  between = sex)
get_anova_table(res.aov) 

#SEX EFFECT
# pwc <- cyto %>%
#   group_by(time) %>%
#   pairwise_t_test(
#     log_il8 ~ sex, paired = FALSE,
#     p.adjust.method = "bonferroni"
#     )
# pwc
 
# EFFECT OF TIME
# pwc2 <- cyto %>%
#   group_by(sex) %>%
#   pairwise_t_test(
#     log_il8 ~ time, paired = TRUE,
#     p.adjust.method = "bonferroni"
#     )
# pwc2

p8 <- ggboxplot(cyto, x = "time", y = "sputum_TNFa", add = "jitter", 
                 color = "sex",
                 palette = c("purple", "turquoise"), 
                 ylab = "TNF alpha (pg/ml)",
                 xlab = FALSE,
                 bxp.errorbar = TRUE)+
  scale_x_discrete(labels= new_labels)+
  theme(legend.title = element_blank())+
  scale_y_continuous(trans = "log10", 
                     breaks = c(1, 10, 100),
                     labels = c("1", "10", "100"))


# pwc2 <- pwc2 %>% add_xy_position(x = "time")
#pwc <- pwc %>% add_xy_position(x = "time")
 
# p8 <- p8 + 
#   stat_pvalue_manual(pwc2, hide.ns = TRUE, step.increase = 0.1,
#                      y.position = 100,
#                      color = "purple")+
#   labs(subtitle = NULL, caption = NULL)
 
# p8 <- p8 +
#   stat_pvalue_manual(pwc, hide.ns = TRUE, step.increase = 0.2,
#                      y.position = 4)+
#   labs(subtitle = NULL, caption = NULL)
p8
```
```{r combine cyto by sex}
combined <- grid.arrange(p5, p6, p7, p8, ncol = 2)
png("clinical_supp_figures/Cyto_bySex.png", width = 13, height = 8, units = "in", res = 300)
grid.draw(combined)
grid.text("A", x = unit(0.01, "npc"), y = unit(0.95, "npc"), just = "left", gp = gpar(fontsize = 14, fontface = "bold"))
grid.text("B", x = unit(0.5, "npc"), y = unit(0.95, "npc"), just = "left", gp = gpar(fontsize = 14, fontface = "bold"))
grid.text("C", x = unit(0.01, "npc"), y = unit(0.48, "npc"), just = "left", gp = gpar(fontsize = 14, fontface = "bold"))
grid.text("D", x = unit(0.5, "npc"), y = unit(0.48, "npc"), just = "left", gp = gpar(fontsize = 14, fontface = "bold"))

dev.off()
```




